Computational Prediction of Boundaries of 3D Genomic Domains in Class Imbalance Settings

### 2019-01-17 Manuscript/manuscript.Rmd
manuscript edits

### FEATURE ENGINEERING USING GENOMIC ANNOTATIONS

@@

To assess the predictive performance for each predictor type, we performed individual random forest models utilizing each specific predictor type, in addition to a model with all combined types. Precision-Recall (PR) Curves, and subsequent area under the PR curves (AUPRC), were used to evaluate which predictor type contributed to the best model performance.

### VARIABLE REDUCTION

@@

4. **SMOTE** (Synthetic Minority Over-Sampling Technique) [@chawla2002smote]: This method incorporated both random under- and over-sampling. Under-sampling is performed without replacement from the majority class, while over-sampling is performed by creating new synthetic observations using the minority class. To account for the bias introduced by the under-sampling, we performed 50-iterations of SMOTE The user is able to specify the parameters denoting the percent of over- and under-sampling from the two classes. We chose the parameters that would create perfectly balanced classes. The `DMwR` R package (version 3.4.2) was used to perform SMOTE.

@@

A classification algorithm in the form of a random forest was performed and validated using the same test set for all models in the ensemble (Figure 3 F). 10-fold cross validation was used to reduced bias due to random dataset generation. The default number of features to consider at each node of the random forest algorithm was set as the square root of the number of features in the model. Likewise, the number of ensemble trees to aggregate was set at 500. Once the model was implemented, it was assessed using the test set, and performance metrics were then recorded.

@@

We evaluated and compared Hi-C datasets at four different resolutions R = {10 kb, 25 kb, 50 kb, 100 kb}, used three predictor types P = {OC, OP, Distance}, two variable reduction techniques V = {LASSO, Elastic-Net}, and four class balancing techniques C = {None, RUS, ROS, SMOTE}. Therefore, a system produced from the ensemble framework is defined as: E = {r, p, v, c}, where r $\in$ R, p $\in$ P, v $\in$ V, and c $\in$ C. Letting |X| denote the cardinality of each set, we evaluated |R| x |P| x |V| x |C| different ensemble systems, totaling 96 (=4 $\times$ 3 $\times$ 2 $\times$ 4) separate models. For simplicity, we present the most significant results, with the complete results available as Supplementary Tables.

@@

The coordinates of the corresponding 10 kb resolution contact maps resulted in a total of 44948 genomic bins [??? To be modified for the whole genome] (Table 2). There were only 3083 (6.9%) bins that overlapped with a flanked TAD boundary [??? What is flanked TAD boundary? Why do we care?]. An increase in resolution resulted in an overall decrease in the number of genomic bins, as well as a slighly less degree of class imbalance. The percentage of imbalance for each of the other resolutions were found to be  14.0%, 13.7%, and 12.2% for 25 kb, 50kb, and 100kb resolutions respectively. [??? Add exact numbers of class imbalance for each resolution; !!! added exact numbers] (Supplementaty Table 2). [??? Do not understand. I understand the ratio of #minority/#majority, but what % tells?; !!!was overally average of class imbalance for the 4 data resolutions; removed sentence; doesn't seem relevant ??? We should rewrite using the ratio of #minority/#majority].  After applying each resampling technique to the training sets for each data resolution, we obtained perfectly balanced classes (Figure 4). That is, the ratio of bins that overlapped with flanked TAD boundaries to bins that did not was 1:1.

### CLASS IMBALANCE HINDERS PERFORMANCE

We first assessed performances on random forest models built on fully unbalanced training sets across each resolution, for each predictor type (Figure 5). When looking at the the conventional metrics used for assessment (accuracy and AUC), models that were not corrected for class imbalance appeared to perform reasonably well.  However, when we considered the F1-score and MCC, we observed a significant degradation in performance across both resolution and predictor type. This was due in part to each models' inability to correctly classify the genomic bins that overlapped with flanked TAD boundaries (true positives, Supplementary Table 3). Even without correcting for the class imbalance, we found that, within predictor types, models with distance type predictors consistently outperformed the other two predictor types across all performance metrics.

### RANDOM UNDER-SAMPLING IMPROVES MODEL PERFORMANCE

When comparing re-sampling techniques it was found that performing 50 iterations of random under-sampling yielded the best performances for both F1-Scores and MCC metrics, across all resolutions and predictor types (Figure 6). Models using 50-iterations of the SMOTE algorithm performed comparitively similar to those using RUS, across all resolutions and predictor types (supplementary Figure 1A). However, it was observed that models using 50-iterations of random over-sampling performed substantially worse than both RUS and SMOTE (Supplementary Figure 1B). More importantly, there was a notable decrease in both accuracy and AUC metrics when performing random under-sampling compared to models with no class balancing. This was evidenced by the more balanced interplay between the componenets of the confusion matrices created when validating the models (Supplementary Table 4). That is, less weight was carried by the true negative cases. This feature was also observed for models using SMOTE and ROS (Supplementary Table 5). As was the case for models with no class-balancing, there was a clear improvement when performing RUS for models with distance-type predictors, across all resolutions.

### DISTANCE TYPE PREDICTORS OUT-PERFORM OTHER PREDICTOR TYPES

[??? Why not using all predictors together? Combine it together with the previous steps, testing no sampling, RUS, and the effect of distance; !!!running on cluster]

In order to more efficiently assess the impact of feature engineering on predictive performance, we directly compared models using each of the three predictor types, across all resolutions, while using 50 iterations of random under-sampling to account for the class imbalance. It was observed that the models built using distance-type features performed substanially better than the other two (Figure 7). This improvement in performance was observed across each of the 4 performance metrics considered. Interestingly, we found that models with overlap percent (OP) type predictors performed worse than those with simple overlap counts (OC), regardless of data resolution. This is notable because, OP predictors can be thought of as a more refined version of OC predictor. Likewise, there have been established methods from previous research that have used OP predictors when attempting to determine which genomic features are most predictive of TAD boundary formation [@mourad2016computational].

### PERFORMANCES DEGRADE AS RESOLUTION INCREASES [??? Terminology: 5kb - high resolution. 100kb - low resolution. Adjust the text]

Finally, we directly compared the performances of models across the different resolutions. The comparisons consisted of random forest models using distance type predictors, elastic-net regularization, and 50 iterations of random under-sampling. The model using 5kb resolution domain data was found to perform slightly better compared to the 25kb and 50kb resolutions (Figure 9). A sharp decrease in performance was seen in the 100kb resolution. [??? This message is already evident from previous paragraphs and results. Not necessary to repeat it]

The differences in performances was most evident when comparing the variable importances across the 4 resolutions. [??? You never said and justified that you use distance predictor.] We found that for the 5kb resolution, the most predictive elements included insulator proteins, DNase hypersensitive sites, the transcription factor CTCF, as well as histone modifications H3k9ac, H2az, and H3k4me2 (Figure 10 A). Although, the insulator protein is featured as a top predictor in 2 of the other 3 resolutions, the rest of the predictors were less consistent among each other and less in agreement with the literature (Figure 10 B-D). The results from Figures 9 and 10 confirm that the model results appear to degrade as resolution increases with respect to both prediction and the identification of genomic elements associated with TAD boundaries.

>>>>

### FEATURE ENGINEERING AND REDUCTION USING GENOMIC ANNOTATIONS

@@

To assess the predictive performance for each predictor type, we performed individual random forest classification algorithms utilizing each set predictor types as the feature space, in addition to a model with all combined types. Precision-Recall (PR) Curves, and subsequent area under the PR curves (AUPRC), were used to evaluate which predictor type contributed to the best model performance. Recursive feature elimination (RFE) was then used to determine the optimal number of features for models within each set of predictor types. Here, random forest models were used to assign weights to a set features of a given size. The optimal number of features was determined by the AUPRC of the corresponding model.

@@

4. **SMOTE** (Synthetic Minority Over-Sampling Technique): This method incorporates both random under- and over-sampling. Under-sampling is performed without replacement from the majority class, while over-sampling is performed by creating new synthetic observations using the minority class [@chawla2002smote]. To account for the bias introduced by the under-sampling, we performed 50-iterations of SMOTE The user is able to specify the parameters denoting the percent of over- and under-sampling from the two classes. We chose the parameters that would create perfectly balanced classes. The `DMwR` package in R (version 3.4.2) was used to perform SMOTE.

5. **ROSE** (Random Over-Sampling Examples): Similar to SMOTE, ROSE combines combines techniques of over-sampling and under-sampling. However, contrary to SMOTE, all of the data generated by ROSE is synthetic. The data is created by sampling from one of the two classes, then generating a new example in its neighborhood using a smoothed bootstrap approach. The shape of the neighborhood is determined by the kernal density estimate of the two classes [@lunardon2013r]. Another important distinction between ROSE and SMOTE is that, for ROSE, the creation of perfectly balanced classes is unlikely due to the randomness in sampling between the two classes. The `ROSE` package in R (version 3.4.2) was used to perform ROSE.

@@

A classification algorithm in the form of a random forest was performed and validated using the same test set for all models in the ensemble. To reduce bias due to random dataset generation, 10-fold cross validation was used. The default number of features to consider at each node of the random forest algorithm was set as the square root of the number of features in the model. Likewise, the number of ensemble trees to aggregate was set at 500. Once the model was implemented, it was assessed using the test set, and performance metrics were then recorded.

@@

We evaluated and compared Hi-C datasets at four different resolutions R = {10 kb, 25 kb, 50 kb, 100 kb}, used three predictor types P = {OC, OP, Distance}, two variable reduction techniques V = {LASSO, Elastic-Net}, and five class balancing techniques C = {None, RUS, ROS, SMOTE, ROSE}. A system produced from the ensemble framework was defined as: E = {r, p, v, c}, where r $\in$ R, p $\in$ P, v $\in$ V, and c $\in$ C. Letting |X| denote the cardinality of each set, we evaluated |R| x |P| x |V| x |C| different ensemble systems, totaling 120 (=4 $\times$ 3 $\times$ 2 $\times$ 5) separate models. For simplicity, we present the most significant results, with the complete results available as Supplementary Tables.

@@

The coordinates of the corresponding 10 kb resolution contact maps resulted in a total of X genomic bins [??? To be modified for the whole genome]. There were only Y bins that overlapped with a flanked TAD boundary representing the minority class, for an imbalance ratio of (X-Y):Y. After applying each resampling technique to the training set, we obtained perfectly balanced classes for all except the ROSE technique (Figure 4).

An increase in resolution resulted in an overall decrease in the number of genomic bins, as well as a slight decrease in the imbalance ratio (Supplementary Figure 2). After performing the resampling techniques, the resulting imbalance ratios remained similar across the other 3 data resolutions (Supplementary Figure 2).

### DISTANCE-TYPE PREDICTORS PERFORM BEST
